\documentclass{article}
\include{aips2defs}
\usepackage[colorlinks]{hyperref}
\usepackage{epsf}
\setlength{\textwidth}{6.5in}
\setlength{\evensidemargin}{0.0in}
\setlength{\oddsidemargin}{0.0in}
\parskip=5truept
\newcommand{\nc}{$N_c$}
\newcommand{\nr}{$N_r$}
\newcommand{\nf}{$N_f$}
\newcommand{\nfs}{$N_f^*$}
\newcommand{\nl}{$N_l$}
\newcommand{\ncat}{$N_{cat}$}
\newcommand{\defline}[1]{\cline{1-5}
\multicolumn{5}{|l|}{#1} \\
\cline{1-5}}
\newcommand{\definetable}[2]
{
	\vfill\newpage
	\subsection{#1}
        \vspace{0.15in}
        \small
	\begin{tabular}{|l|p{1.25in}|l|p{.9in}|p{1.4in}|}
	\hline
	\multicolumn{5}{|c|}{\bf #1}\\ 
	\cline{1-5}
        \multicolumn{1}{|c|}{Name}&\multicolumn{1}{|c|}{Format}&
        \multicolumn{1}{|c|}{Units}&\multicolumn{1}{|c|}{Measure}&
        \multicolumn{1}{|c|}{Comments}\\
        \cline{1-5}
        #2
        \hline
	\end{tabular}
}
\begin{document}

\title{MeasurementSet definition version 3.0}
\author{eds.}
\date{February, 2018}
\maketitle

\ifpdf
\else
\href{264.pdf}{A pdf version of this note is available.}
\fi

\tableofcontents 
\section{Summary} 

This note describes a revised MeasurementSet definition which extends
that previously given in casa/casacore Note 229(Kemball and Wieringa
2000).  The new definition is designed to support features that were
felt missing from v2.0 or tedious to deal with, provide support for
VLBI processing and non dish arrays.

\section{Introduction}

The MeasurementSet (MS) defines the format in which visibility and
single-dish data are stored for packages using casa/casacore code.


\begin{itemize}

\item{{\bf Explicit keys} One of the major complaints of ms v2.0 is
    the use of row ids in subtables as implicit keys. In v3.0 we are
    moving to explicit keys; e.g ANTENNA subtable will have a
    column ANTENNA\_ID so that when making a subset MS re-indexing is
    not needed in the main table.}


\item{{\bf Single-dish processing:} FLOAT\_DATA is going to be replaced
    by DATA unifying it with interferometer. It is left to the column
    implementation to store floats efficiently rather than have
    multiple columns defined for data types }


\item{{\bf Synthesis calibration:} The need for calibrations table as
    part the MeasurementSet has been felt by several telescopes and
    those using v2.0 (e.g ALMA and EVLA) have been using non standard
    tables to carry calibration type information.}

\item{{\bf Data,weight, flag versioning} Non standard columns (e.g
    CORRECTED\_DATA) or kludges have been used by casa (for e.g) to
    deal with having version of data, flag or weight. These 3 columns
    are having the same shapes in every row of an msv3.0.  Whenever
    changes are made example flagging or calibration and a new version
    is needed. A new triplet of columns will be made. It is left to
    software implementation to do the smart thing (not making
    unnecessary copies). E.g if flagging only is done and need to be
    saved in a new version of FLAG column, the new version of DATA an
    WEIGHT are going to be just virtual columns pointing back to the
    previous version.}

\item{{\bf Data Description}The concept of data description is being
    deprecated it saved a column of integers but users found it
    confusing. SPECTRAL\_WINDOW\_ID and POLARIZATION\_ID are explicit
    columns in the main table}

\item{{\bf VLBI data reduction:} From v2.0 added an optional subtable
    INTERFEROMETER\_MODEL}

\item{{\bf Phased Array interferometers:} A new optional subtable,
    PHASED\_ARRAY, which will be defining which antennas form a phased
    array station.

\end{itemize}



Some features:

\begin{itemize}

\item{{\bf Incremental change:} The changes proposed here are designed
to be as incremental as possible and taking into account what usage at different telescopes and software packages has found lacking.}

\item{{\bf Compatibility:} Some of the changes proposed here are not backwards compatible}

\item{{\bf Calibration information:} Many of the direct and peripheral information stored at data collection can be stored as calibration terms (Jones or Mueller matrices). Therefore MS v3.0 will carry in its definition optional calibration table which may carry monitoring information that can be used to modify the data, weights and flags }

\item{{\bf Storage:} A future document will provide results and suggestion for data storage structures for some commonly used access patterns and storage system in use.}

\item{{\bf Multi measurement sets:} In a different document we will
    describe the format of having an MS which consists of valid sub
    MSs without physical concatenation}


\end{itemize}


\section{Summary of changes}

This section contains a description of the changes proposed for each
table in the MS definition. A full definition of the v3.0 MS format is
given in Section 5.

\subsection{MAIN table}

\begin{itemize}


\item{{\bf removal of FLAG\_CATEGORY}


This column has not been used effectively. With the versioning of FLAG below it is redundant.

}
\item{{\bf Removal of SIGMA}
WEIGHT being always considered to be $\frac{1}{\sigma^2}$ this a redundant column.


\item{{\bf Removal of FLAG\_ROW}
Unnecessary and confusing column when used in conjunction with WEIGHT
}

\item{{\bf Redefinition of WEIGHT to be WEIGHT\_SPECTRUM}
With most of contemporary inteferometers being spectral machines with possibility of distinctive weights per channel, usage of MSv2.0 
WEIGHT is low w.r.t WEIGHT\_SPECTRUM. In this version WEIGHT column will have the same shape as the data.
}

\item{{\bf Versioning of (DATA, WEIGHT, FLAG) triplet} These three columns will have the same shapes. There may be many versions them in an MS. But the active version will be refered to as DATA, WEIGHT and FLAG.  The versions need not be copies. For e.g if only DATA is corrected and new version is made. The active DATA will refer to DATA1 column while the active WEIGHT, FLAG (thus WEIGHT1 and FLAG1) will refer to the to WEIGHT0 and FLAG0 respectively. We live it to the software to implement these in the most optimal fashion without making duplicate copies. 

}

\item{{\bf Merging of DATA and FLOAT\_DATA}
It is unnecessary to have explicit DATA and FLOAT\_DATA. The software implementation can do the optimal storage in the presence of float data only and serve it as complex.
}

\item{{\bf Replacement of DATA\_DESC\_ID} This unique ID has caused more confusion than the extra column of integers is worth. In this version we are reverting to having explicit POLARIZATION\_ID and SPECTRAL\_WINDOW\_ID which refers to keys in the POLARIZATION and SPECTRAL\_WINDOW subtables directly. DATA\_DESC\_ID column is removed.

}


\end{itemize}
\subsection{ANTENNA table}

\item {{\bf ANTENNA\_ID} An explicit ID column is newly defined}.





\subsection{ARRAY table}


\begin{itemize}

\item {{\bf ARRAY\_ID} An explicit ID column is newly defined}.


\item{{\bf ARRAY\_CENTER} A new column which will give the full Measures position of where the array center is assumed for each ARRAY\_ID}



\subsection{FEED table}

It is proposed that the ARRAY\_ID column be removed from the FEED
table, for reasons discussed in Section 3.2.1.  The format of the BEAM
sub-table (indexed on BEAM\_ID) is deferred for later specification.
In addition, it is proposed that the PHASED\_FEED\_ID be made optional.

A new column is proposed to store the focus length, as:

\begin{tabbing}
FOCUS\_LENGTH\quad\quad \= Double\quad\quad \= 
(Focus length) \quad\quad \= (Opt.) \\ 
\end{tabbing}


\subsection{FIELD table}

Changes applicable to the FIELD table are discussed in this section.

\subsubsection{Direction information}

With the definition of a POINTING sub-table (see Section 4.0.1), the
POINTING\_DIR and POINTING\_DIR\_RATE columns can be removed from the
FIELD table. 

It is also proposed that the DELAY\_DIR\_RATE, PHASE\_DIR\_RATE and
REFERENCE\_DIR\_RATE columns be replaced by extending the DELAY\_DIR,
PHASE\_DIR and REFERENCE\_DIR columns to dimension (2, NUM\_POLY+1) to
represent higher order terms. This allows the representation of more
specialized correlation modes. The polynomial order is added as a new
column:

\begin{tabbing}
NUM\_POLY\quad\quad \= Int\quad\quad \= (Polynomial order)
 \quad\quad \= (Reqd.) \\
\end{tabbing}

\subsubsection{Ephemeris identification}

A separate identifier is proposed to identify the ephemeris data
associated with time-variable FIELD position data, as obtained for
moving sources. These data are stored in a separate EPHEMERIS
table. This has the added advantage of identifying multiple revisions
of the ephemeris, which may vary between observation and final
reduction. The proposed new column is:

\begin{tabbing} 
EPHEMERIS\_ID\quad\quad  \= Int\quad\quad\quad\quad \= 
(Ephemeris identifier)\quad\quad\quad\quad \= (Opt.) \\
\end{tabbing}

\subsubsection{Indexing}

It is further proposed that the FIELD\_ID column be removed and that
the table be directly indexed using the FIELD\_ID value in MAIN. A
row flag is proposed to invalidate the entry, as follows:

\begin{tabbing}
FLAG\_ROW\quad\quad  \= Bool\quad\quad \= (Row flag)\quad\quad \= (Reqd.)\\
\end{tabbing}

\subsection{FLAG\_CMD}

As discussed in Section 3.1.8, this table is intended to hold all
global flag commands applied to the MS. Their cumulative effect is
reflected in the MAIN column FLAG\_CATEGORY, in the category FLAG\_CMD,
and therefore further in FLAG and FLAG\_ROW. In this proposal, the
FLAG\_CMD records are represented as character strings, using the MS
column names in a limited syntax to represent flagging operations,
each with associated labeling information:

\begin{tabbing} 
FLAG\_COMMAND\quad\quad\quad\quad \= String\quad\quad \= (Flagging command)
\quad\quad\quad\quad \= (Reqd.) \kill
TIME \> Double \> (Midpoint of interval) \> (Reqd.) \\
INTERVAL \> Double \> (Time interval) \> (Reqd.) \\
TYPE \> String \> (FLAG or UNFLAG) \> (Reqd.) \\
REASON \> String \> (Reason) \> (Reqd.) \\
LEVEL  \> Int    \> (Flag level) \> (Reqd.) \\
SEVERITY \> Int \> (Flag severity) \> (Reqd.) \\
APPLIED \> Bool \> (Flag reflected in MAIN ?) \> (Reqd.) \\
COMMAND \> String \> (Flagging command) \> (Reqd.) \\
\end{tabbing}

\subsection{OBSERVATION table}

It is proposed that the CORR\_SCHEDULE field be removed (see Section
4.0.4), which is accessed via the PROCESSOR sub-table. The telescope
name, which was previously stored in the ARRAY sub-table, is proposed
to be recorded as:

\begin{tabbing} 
TELESCOPE\_NAME\quad\quad \= String\quad\quad \= (Telescope name (eg WSRT))
 \quad\quad \= (Reqd.) \\
\end{tabbing}

The observation date is required as:

\begin{tabbing} 
TIME\_RANGE\quad\quad \= Double(2)\quad\quad \= 
(Start and stop times of observation)
 \quad\quad \= (Reqd.) \\
\end{tabbing}

A new column is proposed to identify the type of the unmodified
schedule summary is SCHEDULE, namely:

\begin{tabbing} 
SCHEDULE\_TYPE\quad\quad \= String\quad\quad \= 
(Schedule type (eg. VEX)) \quad\quad \= (Reqd.) \\
\end{tabbing}

A field is proposed to store the log file recorded during observations, as:

\begin{tabbing}
LOG\quad\quad \= String(*)\quad\quad \= (Observing log)
 \quad\quad \= (Opt.) \\
\end{tabbing}

A field is proposed to store the target release date for the project:

\begin{tabbing}
RELEASE\_DATE\quad\quad \= Double\quad\quad \= (Release date)
 \quad\quad \= (Opt.) \\
\end{tabbing}

A row flag is proposed to invalidate the entry as:

\begin{tabbing}
FLAG\_ROW\quad\quad \= Bool\quad\quad \= (Row flag)\quad\quad \= (Reqd.)\\
\end{tabbing}

\subsection{OBS\_LOG table}

It is proposed that this table be renamed as HISTORY, and that its
primary function be to act as a processing log. New columns
are proposed as follows:

\begin{tabbing}
PRIORITY\quad\quad\quad \= String\quad\quad \= 
(Message priority (eg. DEBUGGING, WARN, NORMAL or SEVERE)
 \quad\quad \= (Reqd.) \\
ORIGIN             \> String  \> 
(Source code location from which message originates) \> (Reqd.) \\
OBJECT\_ID         \> String  \>
(Originating ObjectID, if available) \> (Reqd.) \\
APPLICATION        \> String \>
(Application name) \> (Reqd.) \\
\end{tabbing}

User notes will be identified by a specific ORIGIN string.

This is also a useful table to store a record of the Glish commands
and parameters used to invoke the application. This allows parts of
the data reduction to be repeated with minimal effort. An automatic
recording mechanism needs to be invoked by individual methods or
applications for this purpose. This information can be represented in
a string format as:

\begin{tabbing}
CLI\_COMMAND\quad\quad \= String(*)\quad\quad \= 
(Glish parameter or command sequence)
 \quad\quad \= (Reqd.) \\
APP\_PARAMS \> String(*) \> (Application parameters) \> (Reqd.)
\end{tabbing}

 It is recognized that a mechanism to log all ad hoc CLI commands
affecting a MeasurementSet is impractical, but this field should be
filled by all standard applications.

\subsection{SOURCE table}

The changes proposed for the SOURCE table are given below:

\subsubsection{Source model}

Information on the detailed source model needs to be specified at
several points in calibration and imaging, and is properly selected at
that time. A field is added here to record the assignment of a
component source model for each SOURCE\_ID, as this is made during
calibration. This assigned source model will never be used in an
application without explicit notification or confirmation, and is not
intended as a default source model.

\begin{tabbing}
SOURCE\_MODEL\quad\quad \= TableRecord\quad\quad \= 
(Assigned CSM) \quad\quad \= (Opt.) \\
\end{tabbing}

\subsubsection{Spectral line information}

Information on the systemic velocity and rest frequency of spectral
lines observed towards individual sources is required in various
stages of line calibration, particularly for Doppler tracking of
individual transitions in software. The indices SOURCE\_ID and
SPECTRAL\_WINDOW\_ID are a good composite key for this information and
it is proposed that they be added to the SOURCE table. A general
extension for multiple spectral lines per spectral window is proposed
as follows:

\begin{tabbing}
REST\_FREQUENCY\quad\quad \= Double(NUM\_LINES)\quad\quad \=
(Rest frequency for this transition)\quad\quad\quad\quad\quad \= (Reqd.) \kill
NUM\_LINES \> Int \> 
(No. of transitions for this spectral window) \> (Reqd.) \\
TRANSITION           \> String(NUM\_LINES)  \> 
(Transition name) \> (Reqd.) \\
REST\_FREQUENCY       \> Double(NUM\_LINES) \>
(Rest frequency for this transition) \> (Reqd.) \\
SYSVEL               \> Double(NUM\_LINES) \>
(Systemic velocity) \> (Reqd.) \\
\end{tabbing}

\subsubsection{Pulsar information}

 The PULSAR\_ID is moved from the MAIN table to the SOURCE table, as
proposed in Section 3.1.6. This points to the PULSAR table, containing
pulsar source information.

\subsection{SPECTRAL\_WINDOW table}

As discussed in Section 3.2.1, it is proposed that the ARRAY\_ID key be
removed.  In addition, the polarization labeling data CORR\_TYPE,
CORR\_PRODUCT and NUM\_COR are moved to the POLARIZATION sub-table, as
discussed in Section 3.1.3.

\subsubsection{Name}

A spectral window name, is proposed as:

\begin{tabbing}
NAME\quad\quad \= String\quad\quad \= 
(Spectral window name) \quad\quad\quad \= (Reqd.) \\
\end{tabbing}


\subsubsection{Frequency groups}

It is important to identify which SPECTRAL\_WINDOW\_ID's are
associated for calibration purposes as part of a broader frequency
group.  This is necessary in certain calibration applications,
including multi-band VLBI fringe-fitting. Two new columns are
added for this purpose:

\begin{tabbing}
FREQ\_GROUP\_NAME\quad\quad \= String\quad\quad \=
(Frequency group name)\quad\quad \= (Reqd.) \kill
FREQ\_GROUP \> Int \> (Frequency group) \> (Reqd.) \\
FREQ\_GROUP\_NAME \> String \> (Frequency group name) \> (Reqd.) \\
\end{tabbing}

\subsubsection{Doppler tracking}

The REST\_FREQUENCY column is replaced by a DOPPLER\_ID field
reflecting the Doppler tracking parameters used for the spectral
window. These data are stored in a separate DOPPLER table which
references the line parameters defined in Section 3.9.2.

\begin{tabbing}
DOPPLER\_ID\quad\quad \= Int\quad\quad \= 
(Doppler identifier) \quad\quad\quad\quad\quad \= (Opt.) \\
\end{tabbing}

\subsubsection{Sideband and baseband-converter information}

Two new spectral window attributes are added to record baseband
converter and sideband information:

\begin{tabbing}
NET\_SIDEBAND\quad\quad \= Int\quad\quad \= 
(Upper or lower) \quad\quad\quad\quad\quad \= (Reqd.) \\
BBC\_NO           \> Int  \> 
(Baseband converter no.) \> (Opt.) \\
BBC\_SIDEBAND     \> Int \>
(BBC sideband) \> (Opt.) \\
\end{tabbing}

\subsubsection{Receiver information}

An additional index is proposed to reference receiver information:

\begin{tabbing}
RECEIVER\_ID\quad\quad \= Int\quad\quad \= 
(Receiver identifier) \quad\quad\quad\quad\quad \= (Opt.) \\
\end{tabbing}

This points to an optional RECEIVER sub-table.

\subsubsection{Associated spectral windows}

Individual spectral windows may hold a unique relationship to others,
such as that between channel-zero data and the associated, unaveraged
spectrum. Other may have a subset or equivalence relationship, and it
is very useful to record this information. Additional columns for this
purpose are proposed as follows:

\begin{tabbing}
ASSOC\_SPW\_ID\quad\quad \= Int(*)\quad\quad \= 
(Associated spw\_id.) \quad\quad\quad\quad\quad \= (Opt.) \\
ASSOC\_NATURE           \> String(*)  \> 
(Nature of association) \> (Opt.) \\
\end{tabbing}

\subsubsection{Frequency information}

The nominal channel spacing, actual spectral resolution and effective
noise bandwidth are distinct, and it is proposed that they be
decoupled by adding new columns:

\begin{tabbing}
CHAN\_WIDTH\quad\quad \= Double(*)\quad\quad \= 
(Nominal channel width) \quad\quad\quad\quad\quad \= (Reqd.) \\
EFFECTIVE\_BW \> Double(*) \> (Effective bandwidth) \> (Reqd.)\\
\end{tabbing}

The RESOLUTION column then denotes the actual spectral resolution.  A
new column is proposed to allow a row-based Frequency Meaure
reference, as:

\begin{tabbing}
MEAS\_FREQ\_REF\quad\quad \= Int\quad\quad \= (Frequency ref.)\quad\quad 
 \= (Reqd.) \\
\end{tabbing}

\subsubsection{Row flag}

It is proposed that a flag column be added to validate or invalidate
the contents of the row, as follows:

\begin{tabbing}
FLAG\_ROW\quad\quad \= Bool\quad\quad \= (Row flag)\quad\quad \= (Reqd.)\\
\end{tabbing}

\subsection{SYSCAL table}

As discussed in Section 3.2.1 it is proposed that the ARRAY\_ID key be
removed.  Further specific changes to the SYSCAL table are listed
below:

\subsubsection{Number of receptors}

This information is redundant as it is already specified in the FEED
sub-table, and it is proposed that NUM\_RECEPTORS be removed.

\subsubsection{Measured temperatures}

Optional antenna temperature fields are proposed as follows, where
\nr=NUM\_RECEPTORS and \nf=NUM\_CHAN:

\begin{tabbing}
TANT\_TSYS\_SPECTRUM\quad\quad \= Float(\nr,\nf)\quad\quad \=
(${{T_{ant}}\over{T_{sys}}}$ spectrum)
\quad\quad\quad\quad\quad\quad\quad\quad\quad 
\= (Opt.) \kill
TANT \> Float(\nr) \> (Antenna temperature $T_{ant}$) \> (Opt.) \\
TSKY \> Float(\nr) \> (Sky temperature) \> (Opt.) \\
TANT\_SPECTRUM   \> Float(\nr,\nf)  \> 
(Antenna temperature spectrum) \> (Opt.) \\
TSKY\_SPECTRUM \> Float(\nr,\nf) \> (Sky temp. spectrum) \> (Opt.) \\
TANT\_TSYS     \> Float(\nr) \>
(${{T_{ant}}\over{T_{sys}}}$) \> (Opt.) \\
TANT\_TSYS\_SPECTRUM     \> Float(\nr,\nf) \>
(${{T_{ant}}\over{T_{sys}}}$ spectrum) \> (Opt.) \\
TANT\_FLAG     \> Bool \>
(Flag for TANT) \> (Opt.) \\
TSKY\_FLAG \> Bool \> (Flag for TSKY) \> (Opt.) \\
TANT\_TSYS\_FLAG     \> Bool \>
(Flag for TANT\_TSYS) \> (Opt.) \\
\end{tabbing}

 Calibration measurements in bands other than those for which there
are recorded data, are accomodated by creating additional spectral
window identifiers in the SPECTRAL\_WINDOW table, and recording the
data in the same format in the SYSCAL table. This allows simultaneous
wide- and narrow-band calibration data to be recorded in a uniform manner.

 Existing calibration quantities in the SYCAL table, including
PHASE\_DIFF, TCAL, TRX and TSYS, and their associated flags, are made
optional, on an equal footing with the other entries.


\subsubsection{Antenna pointing}

There is a key requirement to know where the antenna and feed are
pointed at the time of the calibration measurement. In synthesis,
these data may be taken while interferometer data are not being
recorded for example, but are nonetheless useful in a priori amplitude
calibration. This information is obtained by lookup in the POINTING
sub-table (see Section 4.0.1).

\subsection{WEATHER table}

As in previous sections it is proposed that the ARRAY\_ID qualifier
for ANTENNA\_ID be removed (see Section 3.2.1). Time- and
direction-dependent ionospheric electron column density data (such as
that derived from GPS receiver data) is proposed to be stored in a
separate TEC sub-table. The existing data columns are made optional,
and associated flags are added:

\begin{tabbing}
IONOS\_ELECTRON\_FLAG\quad\quad \= Bool\quad\quad \= 
(Flag for IONOS\_ELECTRON) \quad\quad \= (Opt.) \\
H2O\_FLAG \> Bool \> (Flag for H2O) \> (Opt.) \\
IONOS\_ELECTRON\_FLAG \> Bool \> (Flag for IONOS\_ELECTRON) \> (Opt.) \\
PRESSURE\_FLAG \> Bool \> (Flag for PRESSURE) \> (Opt.) \\
REL\_HUMIDITY\_FLAG \> Bool \> (Flag for REL\_HUMIDITY) \> (Opt.) \\
TEMPERATURE\_FLAG \> Bool \> (Flag for TEMPERATURE) \> (Opt.) \\
WIND\_DIRECTION\_FLAG \> Bool \> (Flag for WIND\_DIRECTION) \> (Opt.) \\
WIND\_SPEED\_FLAG \> Bool \> (Flag for WIND\_SPEED) \> (Opt.) \\
\end{tabbing}

A new column set is proposed to record the dew point:

\begin{tabbing}
DEW\_POINT\_FLAG\quad\quad   \= Bool\quad\quad  \= 
(Flag for DEW\_POINT)\quad\quad \= (Opt.) \kill
DEW\_POINT \> Float \>  (Dew point temperature) \> (Opt.) \\
DEW\_POINT\_FLAG   \> Bool  \> (Flag for DEW\_POINT) \> (Opt.) \\
\end{tabbing}


\section{New sub-tables}

New sub-tables added to MS v3.0 are included in this Section.

\subsubsection{Multi element station based antenna (PHASED_ARRAY subtable)}

Information about elements that makes a station antenna, combination scheme etc

\subsubsection{SCAN subtable}


\subsubsection{BEAM subtable}

\subsubsection{EPHEMERIDES}
This optional table is explicitly defined and is refered to from the FIELD table via the EPHEMERIS\_ID column.



\subsubsection{Associated tables (CAL\_TABLES)}



\section{MS v3.0 layout}

There is a MAIN table containing a number of data columns and keys
into various subtables. There is at most one of each subtable. The
subtables are stored as keywords of the MS, and all defined sub-tables
are tabulated below.  Optional sub-tables are shown in italics.

\begin{tabular}{|l|p{1.5in}|p{1.6in}|} \hline
\multicolumn{3}{|c|}{\bf Subtables}\\ \cline{1-3}
\multicolumn{1}{|c|}{Table}&\multicolumn{1}{|c|}{Contents}&\multicolumn{1}{|c|}{Keys}\\ \cline{1-3}
ANTENNA& Antenna characteristics & ANTENNA\_ID \\
DATA\_DESCRIPTION & Data description & DATA\_DESC\_ID\\
({\it DOPPLER}) & Doppler tracking & DOPPLER\_ID, SOURCE\_ID \\
FEED  &          Feed characteristics & FEED\_ID, ANTENNA\_ID, TIME, SPECTRAL\_WINDOW\_ID \\
FIELD  &         Field position & FIELD\_ID  \\
FLAG\_CMD  &  Flag commands & TIME \\
({\it FREQ\_OFFSET}) & Frequency offset information & FEED\_ID, ANTENNA$n$, FEED\_ID, TIME, SPECTRAL\_WINDOW\_ID \\
HISTORY  & History information & OBSERVATION\_ID, TIME \\
OBSERVATION  &Observer, Schedule, etc & OBSERVATION\_ID \\ 
POINTING & Pointing information & ANTENNA\_ID, TIME \\
POLARIZATION &Polarization setup & POLARIZATION\_ID \\
PROCESSOR & Processor information & PROCESSOR\_ID  \\
({\it SOURCE})  &        Source information & SOURCE\_ID, SPECTRAL\_WINDOW\_ID, TIME \\
SPECTRAL\_WINDOW  &     Spectral window setups & SPECTRAL\_WINDOW\_ID \\
STATE   & State information & STATE\_ID \\
({\it SYSCAL}) &         System calibration characteristics & FEED\_ID, ANTENNA\_ID, TIME, SPECTRAL\_WINDOW\_ID \\
({\it WEATHER})  &       Weather info for each antenna & ANTENNA\_ID, TIME \\
\hline
\end{tabular}

\par
Note that there are two types of subtables. For the first, simpler
type, the key (ID) is the row number in the subtable. 
Examples are FIELD, SPECTRAL\_WINDOW, OBSERVATION
and PROCESSOR. For the second, the key is a collection of parameters,
usually including TIME. Examples are FEED, SOURCE, SYSCAL, WEATHER.

Note that all optional columns are indicated in italics and in parentheses.

\definetable{MAIN table: Data, coordinates and flags}
{
\defline{\bf Columns}
\defline{\em Keywords}
MS\_VERSION   & Float  & & & MS format version \\
{\it (SORT\_COLUMNS)} & String & & & Sort columns \\  
{\it (SORT\_ORDER)}   & String & & & Sort order \\
\defline{\em Key}
TIME  &         Double  &      s   &     EPOCH    & Integration midpoint \\
{\it (TIME\_EXTRA\_PREC)} & Double & s       &     & extraTIME precision \\
ANTENNA1  &      Int    &          &              & First antenna \\
ANTENNA2  &      Int    &          &              & Second antenna \\
{\it (ANTENNA3)} & Int    &          &              & Third antenna \\
FEED1  &      Int       &          &              & Feed on ANTENNA1 \\
FEED2  &      Int       &          &              & Feed on ANTENNA2 \\
{\it (FEED3)} & Int       &          &              & Feed on ANTENNA3 \\
DATA\_DESC\_ID  &  Int   &          &        & Data desc. id.\\
PROCESSOR\_ID  & Int   &          &        & Processor id.\\
{\it (PHASE\_ID)} & Int &               &         & Phase id.\\
FIELD\_ID  &      Int   &          &        & Field id.\\
\defline{\em Non-key attributes}
INTERVAL  &      Double &       s  &      & Sampling interval \\
EXPOSURE  &      Double &       s  &      & The effective integration time \\
TIME\_CENTROID & Double &       s  &    EPOCH & Time centroid \\
{\it (PULSAR\_BIN)} & Int &          &      & Pulsar bin number \\
{\it (PULSAR\_GATE\_ID)} & Int &         &      & Pulsar gate id. \\   
SCAN\_NUMBER  &   Int   &          &      & Scan number  \\ 
ARRAY\_ID  &      Int   &          &      & Subarray number\\
OBSERVATION\_ID  & Int   &          &     & Observation id.\\
STATE\_ID & Int   &          &              & State id.\\
{\it (BASELINE\_REF)} &  Bool  &     &      & Reference antenna \\
UVW       &   Double(3) &      m   & UVW  & UVW coordinates \\
{\it (UVW2)} &  Double(3) &      m   & UVW  & UVW (baseline 2) \\
}


\definetable{MAIN table: continued}
{
\defline{\em Data}
{\it (DATA)}    &    Complex(\nc, \nf) &  & & Complex visibility
matrix (synthesis arrays)\\
{\it (FLOAT\_DATA)} & Float(\nc, \nf) & & & Float data matrix
(single dish) \\
{\it (VIDEO\_POINT)} & Complex(\nc) & & & Video point \\
{\it (LAG\_DATA)} & Complex(\nc, \nl) & & & Correlation function \\
SIGMA   &    Float(\nc) &  & & Estimated rms noise for single channel \\ 
{\it (SIGMA\_SPECTRUM)}   &    Float(\nc, \nfs) &  & & Estimated rms noise \\ 
WEIGHT   &    Float(\nc) &  & & Weight for whole data matrix \\ 
{\it (WEIGHT\_SPECTRUM)}   &    Float(\nc, \nfs) &  & & Weight for each channel\\ 
\defline{\em Flag information}
FLAG    &    Bool(\nc, \nfs) &  &   &  Cumulative data flags \\ 
FLAG\_CATEGORY & Bool(\nc, \nfs, \ncat) & & & Flag categories \\
FLAG\_ROW &    Bool  &  &   &  The row flag \\ 
}
\begin{description}
\item[Notes:]

\item Note that \nl = number of lags, \nc = number of correlators, 
\nf = number of frequency channels, and \ncat = number of flag categories.

\item[MS\_VERSION] The MeasurementSet format revision number,
expressed as $major\_revision.minor\_revision$. This version is 2.0.

\item[SORT\_COLUMNS] Sort indices, in the form "$index_1,\ index_2\
...$, for the underlying MS. A string containing "NONE" reflects no
sort order.  An example might be SORT\_COLUMNS="TIME ANTENNA1
ANTENNA2", to indicate sorting in in time-baseline order.

\item[SORT\_ORDER] Sort order as either "ASCENDING" or "DESCENDING".

\item[TIME] Mid-point (not centroid) of data interval.

\item[TIME\_EXTRA\_PREC] Extra time precision.

\item[ANTENNA$n$] Antenna number ($\geq 0$), and a direct index into
the ANTENNA sub-table $rownr$. For $n > 2$, triple-product data
are implied.

\item[FEED$n$] Feed number ($\geq 0$). For $n > 2$, triple-product
data are implied.

\item[DATA\_DESC\_ID] Data description identifier ($\geq 0$), and
a direct index into the DATA\_DESCRIPTION sub-table $rownr$.

\item[PROCESSOR\_ID] Processor indentifier ($\geq 0$), and a direct
index into the PROCESSOR sub-table $rownr$.

\item[PHASE\_ID] Switching phase identifier ($\geq 0$)

\item[FIELD\_ID] Field identifier ($\geq 0$).

\item[INTERVAL] Data sampling interval. This is the nominal data
interval and does not include the effects of bad data or partial
integration.

\item[EXPOSURE] Effective data interval, including bad data and
partial averaging.

\item[PULSAR\_BIN] Pulsar bin number for the data record. Pulsar data
may be measured for a limited number of pulse phase bins. The pulse
phase bins are described in the PULSAR sub-table and indexed by this
bin number.

\item[PULSAR\_GATE\_ID] Pulsar gate identifier ($\geq 0)$, and a direct
index into the PULSAR\_GATE sub-table $rownr$. 

\item[SCAN\_NUMBER] Arbitrary scan number to identify data taken in
the same logical scan. Not required to be unique.

\item[ARRAY\_ID] Subarray identifier $(\geq 0)$, which identifies data
in separate subarrays, as defined in Section 3.3.

\item[OBSERVATION\_ID] Observation identifier $(\geq 0)$, which identifies
data from separate observations, as defined in Section 3.3.

\item[STATE\_ID] State identifier $(\geq 0)$, as defined in Section 3.1.5.

\item[BASELINE\_REF] Flag to indicate the original correlator
reference antenna for baseline-based correlators (True for ANTENNA1;
False for ANTENNA2).

\item[UVW] $uvw$ coordinates for the baseline from ANTENNE2 to ANTENNA1,
   i.e. the
   baseline is equal to the difference POSITION2 - POSITION1. The UVW given
   are for the TIME\_CENTROID, and correspond in general to the reference type
   for the PHASE\_DIR of the relevant field. I.e. J2000 if the phase reference
   direction is given in J2000 coordinates. However, any known reference is
   valid.
   Note that the choice of baseline direction and UVW definition ($W$ towards
   source direction; $V$ in plane through source and system's pole; $U$ in
   direction of increasing longitude coordinate) also determines the sign of
   the phase of the recorded data.

\item[UVW2] $uvw$ coordinates for the baseline from ANTENNE3 to ANTENNA1
   (triple-product data only), i.e. the
   baseline is equal to the difference POSITION3 - POSITION1. The UVW given
   are for the TIME\_CENTROID, and correspond in general to the reference type
   for the PHASE\_DIR of the relevant field. I.e. J2000 if the phase reference
   direction is given in J2000 coordinates. However, any known reference is
   valid.
   Note that the choice of baseline direction and UVW definition ($W$ towards
   source direction; $V$ in plane through source and system's pole; $U$ in
   direction of increasing longitude coordinate) also determines the sign of
   the phase of the recorded data.

\item[DATA, FLOAT\_DATA, LAG\_DATA] At least one of these columns
should be present in a given MeasurementSet. In special cases one or
more could be present (e.g., single dish data used in synthesis
imaging or a mix of auto and crosscorrelations on a multi-feed single
dish). If only correlation functions are stored in the MS, then \nfs\
is the maximum number of lags (\nl) specified in the LAG table for
this LAG\_ID.  If both correlation functions and frequency spectra are
stored in the same MS, then \nfs\ is the number of frequency channels,
and the weight information refers to the frequency spectra only. The
units for these columns (eg. 'Jy') specify whether the data are in
flux density units or correlation coefficients.

\item[VIDEO\_POINT] The video point for the spectrum, to allow the
full reverse transform.

\item[SIGMA] The estimated rms noise for a single channel, for
each correlator.

\item[SIGMA\_SPECTRUM] The estimated rms noise for each channel.

\item[WEIGHT] The weight for the whole data matrix for each
correlator, as assigned by the correlator or processor.

\item[WEIGHT\_SPECTRUM] The weight for each channel in the data
matrix, as assigned by the correlator or processor. The weight
spectrum should be used in preference to the WEIGHT, when available.

\item[FLAG] An array of Boolean values with the same shape as DATA
(see the DATA item above) representing the cumulative flags applying
to this data matrix, as specified in FLAG\_CATEGORY. Data are flagged
bad if the FLAG array element is True.

\item[FLAG\_CATEGORY] An array of flag matrices with the same shape as
DATA, but indexed by category. The category identifiers are specified
by a keyword CATEGORY, containing an array of string identifiers,
attached to the FLAG\_CATEGORY column and thus shared by all rows in
the MeasurementSet. The cumulative effect of these flags is reflected
in column FLAG. Data are flagged bad if the FLAG array element is
True. See Section 3.1.8 for further details.

\item[FLAG\_ROW] True if the entire row is flagged.

\end{description}

\definetable{ANTENNA: Antenna characteristics}
{
\defline{\bf Columns}
\defline{\em Data}
NAME &       String &  & &  Antenna name \\
STATION &        String & & &  Station name \\
TYPE  &  String &     &      & Antenna type \\
MOUNT  & String &  & &   Mount type:alt-az, equatorial, X-Y, orbiting, bizarre \\
POSITION &   Double(3) &  m   &     POSITION &   Antenna X,Y,Z phase reference positions \\
OFFSET & Double(3)  &   m &     POSITION & Axes offset of mount to FEED REFERENCE point \\
DISH\_DIAMETER &  Double &  m  &      & Diameter of dish\\
{\it (ORBIT\_ID)}    &    Int & & &  Orbit id. \\
{\it (MEAN\_ORBIT)}  &    Double(6) & & & Mean Keplerian elements\\
{\it (PHASED\_ARRAY\_ID)} & Int   &  & &   Phased array id.\\
\defline{Flag information}
FLAG\_ROW & Bool & & & Row flag\\
}
\begin{description}

\item[Notes:] This sub-table contains the global antenna properties
for each antenna in the MS. It is indexed directly from MAIN via
ANTENNA$n$.

\item[NAME] Antenna name (e.g. "NRAO\_140")

\item[STATION] Station name (e.g. "GREENBANK")

\item[TYPE] Antenna type. Reserved keywords include:
("GROUND-BASED" - conventional antennas; "SPACE-BASED" - orbiting
antennas; "TRACKING-STN" - tracking stations).

\item[MOUNT] Mount type of the antenna. Reserved keywords include:
("EQUATORIAL" - equatorial mount; "ALT-AZ" - azimuth-elevation mount; "X-Y"
- x-y mount; "SPACE-HALCA" - specific orientation model.)

\item[POSITION] In a right-handed frame, X towards the intersection of
the equator and the Greenwich meridian, Z towards the pole. The exact
frame should be specified in the MEASURE\_REFERENCE keyword (ITRF or
WGS84). The reference point is the point on the az or ha axis closest
to the el or dec axis.

\item[OFFSET] Axes offset of mount to feed reference point.

\item[DISH\_DIAMETER] Nominal diameter of dish, as opposed to the
effective diameter.

\item[ORBIT\_ID] Orbit identifier. Index used in ORBIT sub-table if
ANTENNA\_TYPE is "SPACE\_BASED".

\item[MEAN\_ORBIT] Mean Keplerian orbital elements, using the
standard convention (Flatters 1998): 
\begin{itemize}
   \item{{\bf 0:} Semi-major axis of orbit ($a$) in $m$.}
   \item{{\bf 1:} Ellipticity of orbit ($e$).}
   \item{{\bf 2:} Inclination of orbit to the celestial equator ($i$) 
                  in $deg.$}
   \item{{\bf 3:} Right ascension of the ascending node ($\Omega$) in $deg.$}
   \item{{\bf 4:} Argument of perigee ($\omega$) in $deg.$}
   \item{{\bf 5:} Mean anomaly ($M$) in $deg.$}
\end{itemize}

\item[PHASED\_ARRAY\_ID] Phased array identifier. Points to a
PHASED\_ARRAY sub-table which points back to multiple entries in the
ANTENNA sub-table and contains information on how they are combined.

\item[FLAG\_ROW] Boolean flag to indicate the validity of this
entry. Set to True for an invalid row. This does not imply any
flagging of the data in MAIN, but is necessary as the ANTENNA index in
MAIN points directly into the ANTENNA sub-table. Thus FLAG\_ROW can be
used to delete an antenna entry without re-ordering the ANTENNA
indices throughout the MS.

\end{description}

\definetable{DATA\_DESCRIPTION: Data description table}{
\defline{\bf Columns}
\defline{\em Data}
SPECTRAL\_WINDOW\_ID & Int & & & Spectral window id.\\
POLARIZATION\_ID & Int & & & Polarization id.\\
{\it (LAG\_ID)} & Int & & & Lag fn. id.\\
\defline{\em Flags}
FLAG\_ROW & Bool & & & Row flag.\\
}
\begin{description}

\item[Notes:] This table define the shape of the associated DATA array
in MAIN, and in indexed directly by DATA\_DESC\_ID.

\item[SPECTRAL\_WINDOW\_ID] Spectral window identifier.

\item[POLARIZATION\_ID] Polarization identifier ($\geq 0$); direct
index into the POLARIZATION sub-table.

\item[LAG\_ID] Lag function identifier ($\geq 0$), and a direct index
into the LAG sub-table $rownr$.

\item[FLAG\_ROW] True if the row does not contain valid data; does not
imply flagging in MAIN.

\end{description}


\definetable{DOPPLER: Doppler tracking information}{
\defline{\bf Columns}
\defline{\em Key}
DOPPLER\_ID  & Int & & & Doppler tracking id.\\
SOURCE\_ID  & Int & & & Source id.\\
\defline{\em Data}
TRANSITION\_ID & Int & & & Transition id.\\
VELDEF & Double & m/s & Doppler & Velocity definition of Doppler shift.\\
}
\begin{description}

\item[Notes:] This sub-table contains frame information 
for different Doppler tracking modes. It is indexed from the
SPECTRAL\_WINDOW\_ID sub-table (with SOURCE\_ID as a secondary index)
and thus allows the specification of a source-dependent Doppler
tracking reference for each SPECTRAL\_WINDOW. This model allows
multiple possible transitions per source per spectral window, but only
one reference at any given time.

\item[DOPPLER\_ID] Doppler identifier, as used in the SPECTRAL\_WINDOW
sub-table.

\item[SOURCE\_ID] Source identifier (as used in the SOURCE sub-table).

\item[TRANSITION\_ID] This index selects the appropriate line from the
list of transitions stored for each SOURCE\_ID in the SOURCE table.

\item[VELDEF] Velocity definition of the Doppler shift, e.g., RADIO or
OPTICAL velocity in m/s.

\end{description}



\definetable{FEED: Feed characteristics}{
\defline{\bf Columns}
\defline{\em Key}
ANTENNA\_ID &     Int &    &      & Antenna id\\
FEED\_ID &        Int &    &      & Feed id\\
SPECTRAL\_WINDOW\_ID &   Int &    &      & Spectral window id.\\
TIME &   Double  &      s  & EPOCH & Interval midpoint \\
INTERVAL &       Double  &      s     &       & Time interval\\
\defline{\em Data description}
NUM\_RECEPTORS &  Int &    &      & \# receptors on this feed \\
\defline{\em Data}
BEAM\_ID &        Int    &    &      & Beam model\\
BEAM\_OFFSET &    Double(2, NUM\_RECEPTORS)    &     rad & DIRECTION & Beam position offset (on sky but in antenna reference frame).\\
{\it (FOCUS\_LENGTH)} &   Double &    m       &       & Focus length \\
{\it (PHASED\_FEED\_ID)} & Int   &    &      & Phased feed\\
POLARIZATION\_TYPE &       String (NUM\_RECEPTORS) &    &      & Type of polarization to which a given RECEPTOR responds.\\
POL\_RESPONSE &  Complex (NUM\_RECEPTORS, NUM\_RECEPTORS) &    &      & Feed polzn. response\\
POSITION &    Double(3)  & m  &         POSITION & Position of feed relative to feed reference position for this antenna\\
RECEPTOR\_ANGLE &    Double (NUM\_RECEPTORS) &    rad &    & The reference angle for polarization.\\
}
\begin{description}

\item[Notes:] A feed is a collecting element on an antenna, such as a
single horn, that shares joint physical properties and makes sense to
calibrate as a single entity. It is an abstraction of a generic
antenna feed and is considered to have one or more RECEPTORs that
respond to different polarization states. A FEED may have a
time-variable beam and polarization response. 
Feeds are numbered from 0 on each separate antenna for each
SPECTRAL\_WINDOW\_ID. Consequently, FEED\_ID should be non-zero only
in the case of feed arrays, i.e. multiple, simultaneous beams on the
sky at the same frequency and polarization.

\item[ANTENNA\_ID] Antenna number, as indexed from ANTENNA$n$ in
MAIN.

\item[FEED\_ID] Feed identifier, as indexed from FEED$n$ in MAIN.

\item[SPECTRAL\_WINDOW\_ID] Spectral window identifier. A value of -1
indicates the row is valid for all spectral windows.

\item[TIME] Mid-point of time interval for which the feed parameters
in this row are valid. The same Measure reference used for the TIME
column in MAIN must be used.

\item[INTERVAL] Time interval. 

\item[NUM\_RECEPTORS] Number of receptors on this feed. See
POLARIZATION\_TYPE for further information.

\item[BEAM\_ID] Beam identifier. Points to an optional BEAM sub-table
defining the primary beam and polarization response for this FEED. A
value of -1 indicates that no associated beam response is defined.

\item[BEAM\_OFFSET] Beam position offset, as defined on the sky but in
the antenna reference frame.

\item[FOCUS\_LENGTH] Focus length. As defined along the optical axis
of the antenna.

\item[PHASED\_FEED\_ID] Phased feed identifier. Points to a
PHASED\_FEED sub-table which in turn points back to multiple entries
in the FEED table, and specifies the manner in which they are
combined. 

\item[POLARIZATION\_TYPE] Polarization type to which each receptor
responds (e.g. "R","L","X" or "Y"). This is the receptor polarization
type as recorded in the final correlated data (e.g. "RR"); i.e. as
measured after all polarization combiners.

\item[POL\_RESPONSE] Polarization response at the center of the beam
for this feed. Expressed in a linearly polarized basis $({\bf \vec
e_x}, {\bf \vec e_y})$ using the IEEE convention.

\item[POSITION] Offset of feed relative to the feed reference position
for this antenna (see ANTENNA sub-table).

\item[RECEPTOR\_ANGLE] Polarization reference angle. Converts into
parallactic angle in the sky domain.

\end{description}

\definetable{FIELD: Field positions for each source}{
\defline{\bf Columns}
\defline{\em Key}
\defline{\em Data}  
NAME &       String &    &      & Name of field \\
CODE &       String &    &      & Special characteristics of field \\
TIME &   Double  &      s  & EPOCH & Time origin for the directions and rates\\
NUM\_POLY & Int & & & Series order \\
DELAY\_DIR    &      Double(2, NUM\_POLY+1)  &      rad   &     DIRECTION & Direction of delay center. \\
PHASE\_DIR    &      Double(2, NUM\_POLY+1)  &      rad   &     DIRECTION & Phase center. \\
REFERENCE\_DIR &      Double(2, NUM\_POLY+1)  &      rad   &     DIRECTION & Reference center \\
SOURCE\_ID &      Int    &    &      & Index in Source table\\
{\it (EPHEMERIS\_ID)} &   Int    &    &      & Ephemeris id.\\
\defline{Flags}
FLAG\_ROW & Bool & & & Row flag\\
}
\begin{description}

\item[Notes:] The FIELD table defines a field position on the sky. For
interferometers, this is the correlated field position. For single dishes,
this is the nominal pointing direction.

\item[NAME] Field name; user specified.

\item[CODE] Field code indicating special characteristics of the
field; user specified.

\item[TIME] Time reference for the directions and rates. Required to 
use the same TIME Measure reference as in MAIN.

\item[NUM\_POLY] Series order for the *\_DIR columns.

\item[DELAY\_DIR] Direction of delay center; can be expressed as a
polynomial in time. Final result converted to the defined Direction
Measure type.

\item[PHASE\_DIR] Direction of phase center; can be expressed as a
polynomial in time. Final result converted to the defined Direction
Measure type.

\item[REFERENCE\_DIR] Reference center; can be expressed as a
polynomial in time. Final result converted to the defined Direction
Measure type. Used in single-dish to record the associated reference
direction if position-switching has already been applied. For
interferometric data, this is the original correlated field center,
and may equal DELAY\_DIR or PHASE\_DIR.

\item[SOURCE\_ID] Points to an entry in the optional SOURCE subtable, a
value of $-1$ indicates there is no corresponding source defined.

\item[EPHEMERIS\_ID] Points to an entry in the EPHEMERIS sub-table,
which defines the ephemeris used to compute the field position. Useful
for moving, near-field objects, where the ephemeris may be revised over time.

\item[FLAG\_ROW] True if data in this row are invalid, else False. Does not
imply flagging in MAIN.

\end{description}

\definetable{FLAG\_CMD: Flag commands}{
\defline{\bf Columns}
\defline{\em Key}
TIME      &    Double   &   s  &  EPOCH & Mid-point of interval \\
INTERVAL  &    Double   &   s  &        & Time interval \\
\defline{\em Data}  
TYPE      &    String   &      &        & FLAG or UNFLAG\\
REASON    &    String   &      &        & Flag reason\\
LEVEL     &    Int      &      &        & Flag level\\
SEVERITY  &    Int      &      &        & Severity code\\
APPLIED   &    Bool     &      &        & True if applied in MAIN\\
COMMAND   & String  &      &        & Flag command\\
}
\begin{description}

\item[Notes:] The FLAG\_CMD sub-table defines global flagging commands
which apply to the data in MAIN, as described in Section 3.1.8.

\item[TIME] Mid-point of the time interval to which this flagging
command applies. Required to use the same TIME Measure reference as
used in MAIN.

\item[INTERVAL] Time interval.

\item[TYPE] Type of flag command, representing either a flagging
("FLAG") or un-flagging ("UNFLAG") operation.

\item[REASON] Flag reason; user specified.

\item[LEVEL] Flag level $(\geq 0)$; reflects different revisions of
flags which have the same REASON.

\item[SEVERITY] Severity code for the flag, on a scale of 0-10 in
order of increasing severity; user specified.

\item[APPLIED] True if this flag has been applied to MAIN, and update
in FLAG\_CATEGORY and FLAG. False if this flag has not been applied to
MAIN.

\item[COMMAND] Global flag command, expressed in the standard
syntax for data selection, as adopted within the project as a whole.

\end{description}

\definetable{FREQ\_OFFSET: Frequency offset information}{
\defline{\bf Columns}
\defline{\em Key}
ANTENNA1  &   Int  & & & Antenna 1.\\
ANTENNA2  &   Int  & & & Antenna 2.\\
FEED\_ID     &   Int  & & & Feed id.\\
SPECTRAL\_WINDOW\_ID & Int & & & Spectral window id.\\
TIME & Double & s & EPOCH & Interval midpoint\\
INTERVAL & Double & s & & Time interval\\
\defline{\em Data}
OFFSET   &  Double & Hz &  & Frequency offset\\
}
\begin{description}

\item[Notes:] The table contains frequency offset information, to be
added directly to the defined frequency labeling in the
SPECTRAL\_WINDOW sub-table as a Measure offset. This allows bands with
small, time-variable, ad hoc frequency offsets to be labeled as the same
SPECTRAL\_WINDOW\_ID, and calibrated together if required.

\item[ANTENNA$n$] Antenna identifier, as indexed from ANTENNA$n$ in MAIN.

\item[FEED\_ID] Antenna identifier, as indexed from FEED$n$ in MAIN.

\item[SPECTRAL\_WINDOW\_ID] Spectral window identifier.

\item[TIME] Mid-point of the time interval for which this offset is valid.
Required to use the same TIME Measure reference as used in MAIN.

\item[INTERVAL] Time interval.

\item[OFFSET] Frequency offset to be added to the frequency axis for
this spectral window, as defined in the SPECTRAL\_WINDOW
sub-table. Required to have the same Frequency Measure reference as
CHAN\_FREQ in that table.

\end{description}

\definetable{HISTORY: History information}{
\defline{\bf Columns}
\defline{\em Key}
TIME & Double & s &              EPOCH  & Time-stamp for message\\
OBSERVATION\_ID  & Int   &          &        & Points to OBSERVATION table\\
\defline{\em Data}  
MESSAGE  & String &    &      & Log message\\
PRIORITY & String & & & Message priority\\
ORIGIN   & String & & & Code origin\\
OBJECT\_ID & String & & & Originating ObjectID \\
APPLICATION & String & & & Application name\\
CLI\_COMMAND & String(*) & & & CLI command sequence \\
APP\_PARAMS   & String(*) & & & Application paramters\\
}
\begin{description}

\item[Notes:] This sub-table contains associated history information
for the MS.

\item[TIME] Time-stamp for the history record. Required to have the
same TIME Measure reference as used in MAIN.

\item[OBSERVATION\_ID] Observation identifier (see the OBSERVATION table)

\item[MESSAGE] Log message.

\item[PRIORITY] Message priority, with allowed types: ("DEBUGGING",
"WARN", "NORMAL", or "SEVERE").

\item[ORIGIN] Source code origin from which message originated.

\item[OBJECT\_ID] Originating ObjectID, if available, else blank.

\item[APPLICATION] Application name.

\item[CLI\_COMMAND] CLI command sequence invoking the application.

\item[APP\_PARAMS] Application parameter values, in the adopted
project-wide format.
\end{description}

\definetable{OBSERVATION: Observation information}{
\defline{\bf Columns}
\defline{\em Data}  
TELESCOPE\_NAME &      String &    &      & Telescope name\\
TIME\_RANGE       &     Double(2) & s & EPOCH & Start, end times\\
OBSERVER &       String &    &      & Name of observer(s)\\
LOG &   String(*) & & & Observing log \\
SCHEDULE\_TYPE &  String & & & Schedule type\\
SCHEDULE &       String(*) &    &      & Project schedule\\
PROJECT &               String &    &      & Project identification string.\\
RELEASE\_DATE &         Double &  s  & EPOCH & Target release date\\
\defline{\em Flags}
FLAG\_ROW  & Bool & & & Row flag.\\
}

\begin{description}

\item[Notes:] This table contains information specifying the observing
instrument or epoch. See the discussion in Section 3.3 for details. It
is indexed directly from MAIN via OBSERVATION\_ID.

\item[TELESCOPE\_NAME] Telescope name (e.g. "WSRT" or "VLBA").

\item[TIME\_RANGE] The start and end times of the overall observing
period spanned by the actual recorded data in MAIN. Required to use
the same TIME Measure reference as in MAIN.

\item[OBSERVER] The name(s) of the observer(s).

\item[LOG] The observing log, as supplied by the telescope or instrument.

\item[SCHEDULE\_TYPE] The schedule type, with current reserved types
("VLBA-CRD", "VEX", "WSRT", "ATNF").

\item[SCHEDULE] Unmodified schedule file, of the type specified, and
as used by the instrument.

\item[PROJECT] Project code (e.g. "BD46")

\item[RELEASE\_DATE] Project release date. This is the date on which
the data may become public.

\item[FLAG\_ROW] Row flag. True if data in this row is invalid, but
does not imply any flagging in MAIN.
\end{description}

\definetable{POINTING: Antenna pointing information}{
\defline{\bf Columns}
\defline{\em Key}
ANTENNA\_ID  &  Int & & & Antenna id.\\
TIME         &  Double & s & EPOCH & Interval midpoint\\
INTERVAL     &  Double & s &       & Time interval\\
\defline{\em Data}
NAME     &   String & & & Pointing position desc.\\
NUM\_POLY & Int & & & Series order \\
TIME\_ORIGIN & Double & s & EPOCH & Origin for the polynomial\\
DIRECTION & Double(2, NUM\_POLY+1) & rad & DIRECTION & 
  Antenna pointing direction\\
TARGET & Double(2, NUM\_POLY+1) & rad & DIRECTION & Target direction \\
{\it (POINTING\_OFFSET)} & Double(2, NUM\_POLY+1) & rad & DIRECTION &
  A priori pointing correction \\
{\it (SOURCE\_OFFSET)} & Double(2, NUM\_POLY+1) & rad & DIRECTION & Offset from source\\
{\it (ENCODER)} & Double(2) & rad & DIRECTION & Encoder values\\
{\it (POINTING\_MODEL\_ID)} & Int & & & Pointing model id. \\
TRACKING & Bool & & & True if on-position \\
{\it(ON\_SOURCE)} & Bool & & & True if on-source\\
{\it(OVER\_THE\_TOP)} & Bool & & & True if over the top\\
}
\begin{description}

\item[Notes:] This table contains information concerning the primary
pointing direction of each antenna as a function of time. Note that
the pointing offsets for inidividual feeds on a given antenna are
specified in the FEED sub-table with respect to this pointing direction.

\item[ANTENNA\_ID] Antenna identifier, as specified by ANTENNA$n$ in MAIN.

\item[TIME] Mid-point of the time interval for which the information
in this row is valid. Required to use the same TIME Measure reference
as in MAIN.

\item[INTERVAL] Time interval.

\item[NAME] Pointing direction name; user specified.

\item[NUM\_POLY] Series order for the polynomial expressions in
DIRECTION and POINTING\_OFFSET.

\item[TIME\_ORIGIN] Time origin for the polynomial expansions.

\item[DIRECTION] Antenna pointing direction, optionally expressed as
polynomial coefficients. The final result is interpreted as a
Direction Measure using the specified Measure reference. 

\item[TARGET] Target pointing direction, optionally expressed as
polynomial coefficients. The final result is interpreted as a
Direction Measure using the specified Measure reference. This is the
true expected position of the source, including all coordinate
corrections such as precession, nutation etc.

\item[POINTING\_OFFSET] The a priori pointing corrections applied by
the telescope in pointing to the DIRECTION position, optionally
expressed as polynomial coefficients. The final result is interpreted
as a Direction Measure using the specified Measure reference.

\item[SOURCE\_OFFSET] The commanded offset from the source position,
if offset pointing is being used.

\item[ENCODER] The current encoder values on the primary axes of the
mount type for the antenna, expressed as a Direction Measure.

\item[TRACKING] True if tracking the nominal pointing position.

\item[ON-SOURCE] True if the nominal pointing direction coincides with
the source, i.e. offset-pointing is not being used.

\item[OVER-THE-TOP] True if the antenna was driven to this position
"over the top" (az-el mount).
\end{description}

\definetable{POLARIZATION: Polarization setup information}{
\defline{\bf Columns}
\defline{\em Data description columns}
NUM\_CORR   &     Int  &           &        & \# correlations\\
\defline{\em Data}  
CORR\_TYPE & Int(NUM\_CORR) & & & Polarization of correlation \\
CORR\_PRODUCT &  Int(2, NUM\_CORR) &  & & Receptor cross-products \\  
\defline{\em Flags}
FLAG\_ROW & Bool & & & Row flag\\
}
\begin{description}

\item[Notes:] This table defines the polarization labeling of the DATA
array in MAIN, and is directly indexed from the DATA\_DESCRIPTION
table via POLARIZATION\_ID.

\item[NUM\_CORR] The number of correlation polarization products. For
example, for (RR) this value would be 1, for (RR, LL) it would be 2,
and for (XX,YY,XY,YX) it would be 4, etc.

\item[CORR\_TYPE] An integer for each correlation product indicating
the Stokes type as defined in the Stokes class enumeration.

\item[CORR\_PRODUCT] Pair of integers for each correlation product,
specifying the receptors from which the signal originated.  The
receptor polarization is defined in the POLARIZATION\_TYPE column in
the FEED table. An example would be (0,0), (0,1), (1,0), (1,1) to
specify all correlations between two receptors.

\item[FLAG\_ROW] Row flag. True is the data in this row are not valid,
but does not imply the flagging of any DATA in MAIN.

\end{description}

\definetable{PROCESSOR: Processor information}{
\defline{\bf Columns}
\defline{\em Data}
TYPE            & String & & & Processor type \\
SUB\_TYPE & String & & & Processor sub-type\\
TYPE\_ID  & Int & & & Processor type id.\\
MODE\_ID  & Int & & & Processor mode id.\\
{\it (PASS\_ID)}  & Int & & & Processor pass number\\
\defline{\em Flags}
FLAG\_ROW & Bool & & & Row flag\\
}
\begin{description}

\item[Notes:] This table holds summary information for the back-end
processing device used to generate the basic data in the MAIN
table. Such devices include correlators, radiometers, spectrometers,
pulsar-timers, amongst others. See Section 4.0.4 for further details.

\item[TYPE] Processor type; reserved keywords include ("CORRELATOR" -
interferometric correlator; "SPECTROMETER" - single-dish correlator;
"RADIOMETER" - generic detector/integrator; "PULSAR-TIMER" - pulsar
timing device).

\item[SUB\_TYPE] Processor sub-type, e.g. "GBT" or "JIVE".

\item[TYPE\_ID] Index used in a specialized sub-table named as
$subtype\_type$, which contains time-independent processor information
applicable to the current data record (e.g. a JIVE\_CORRELATOR
sub-table).  Time-dependent information for each device family is
contained in other tables, dependent on the device type.

\item[MODE\_ID] Index used in a specialized sub-table named as
$subtype\_type\_mode$, containing information on the processor mode
applicable to the current data record. (e.g. a GBT\_SPECTROMETER\_MODE
sub-table).

\item[PASS\_ID] Pass identifier; this is used to distinguish data records
produced by multiple passes through the same device, where this is
possible (e.g. VLBI correlators). Used as an index into the associated
table containing pass information.

\item[FLAG\_ROW] Row flag. True if data in the row is not valid, but
does not imply flagging in MAIN.

\end{description}

\definetable{SOURCE: Source information}{
\defline{\bf Columns}
\defline{\em Key}
SOURCE\_ID &      Int &    &      & Source id\\
TIME &   Double & s  &   EPOCH  & Midpoint of time for which this set of parameters is accurate\\
INTERVAL &       Double & s &               & Interval\\
SPECTRAL\_WINDOW\_ID &   Int &    &      & Spectral Window id\\
\defline{\em Data description}
NUM\_LINES &  Int & & & Number of spectral lines\\
\defline{\em Data}  
NAME &       String &    &      & Name of source as given during observations\\
CALIBRATION\_GROUP &      Int &    &      & \# grouping for calibration purpose\\
CODE &       String &    &      & Special characteristics of source, e.g. Bandpass calibrator\\
DIRECTION &       Double(2) &  rad  &    DIRECTION &        Direction (e.g. RA, DEC) \\
{\it (POSITION)}  & Double(3) &   m  & POSITION & Position (e.g. for solar system objects) \\
PROPER\_MOTION &  Double(2) &  rad/s & &  Proper motion \\
{\it (TRANSITION)} & String(NUM\_LINES) & & & Transition name\\
{\it (REST\_FREQUENCY)} & Double(NUM\_LINES) & Hz & FREQUENCY & Line rest frequency\\
{\it (SYSVEL)}   & Double(NUM\_LINES)  &  m/s  & RADIAL VELOCITY & Systemic velocity at reference \\
{\it (SOURCE\_MODEL)} & TableRecord & & & Default csm \\
{\it (PULSAR\_ID)} & Int & & & Pulsar id. \\
}
\begin{description}

\item[Notes:] This table contains time-variable source information,
optionally associated with a given FIELD\_ID.

\item[SOURCE\_ID] Source identifier ($\geq 0)$, as specified in the
FIELD sub-table.

\item[TIME] Mid-point of the time interval for which the data in this
row is valid. Required to use the same TIME Measure reference as in
MAIN.

\item[INTERVAL] Time interval.

\item[SPECTRAL\_WINDOW\_ID] Spectral window identifier. A -1 indicates
that the row is valid for all spectral windows.

\item[NUM\_LINES] Number of spectral line transitions associated with
this source and spectral window id. combination.

\item[NAME] Source name; user specified.

\item[CALIBRATION\_GROUP] Calibration group number to which this source
belongs; user specified.

\item[CODE] Source code, used to describe any special characteristics
f the source, such as the nature of a calibrator. Reserved keyword,
including ("BANDPASS CAL").

\item[DIRECTION] Source direction at this TIME.

\item[POSITION] Source position $(x,y,z)$ at this TIME (for near-field
objects).

\item[PROPER\_MOTION] Source proper motion at this TIME.

\item[TRANSITION] Transition names applicable for this spectral window
(e.g. "v=1, J=1-0, SiO").

\item[REST\_FREQUENCY] Rest frequencies for the transitions.

\item[SYSVEL] Systemic velocity for each transition.

\item[SOURCE\_MODEL] Reference to an assigned component source model
table.

\item[PULSAR\_ID] An index used in the PULSAR sub-table to define
further pulsar-specific properties if the source is a pulsar.

\end{description}

\definetable{SPECTRAL\_WINDOW: Spectral window description}{
\defline{\bf Columns}
\defline{\em Data description columns}
NUM\_CHAN   &     Int  &           &        & \# spectral channels \\
\defline{\em Data}  
NAME & String & & & Spectral window name\\
REF\_FREQUENCY     &   Double       &               Hz  &   FREQUENCY&        The reference frequency.\\
CHAN\_FREQ  &   Double(NUM\_CHAN)    &           Hz    &    FREQUENCY& Center frequencies for each channel in the data matrix.\\
CHAN\_WIDTH &   Double(NUM\_CHAN)    &           Hz    &    & Channel width for each channel in the data matrix.\\
MEAS\_FREQ\_REF & Int      & & & FREQUENCY Measure ref.\\
EFFECTIVE\_BW   &     Double(NUM\_CHAN)         &      Hz      &  &
The effective noise bandwidth of each spectral channel\\
RESOLUTION   &     Double(NUM\_CHAN)         &      Hz      &  &
The effective spectral resolution of each channel\\
TOTAL\_BANDWIDTH & Double & Hz &  & total bandwidth for this window \\
NET\_SIDEBAND & Int & & & Net sideband\\
{\it (BBC\_NO)} & Int & & & Baseband converter no.\\
{\it (BBC\_SIDEBAND)} & Int & & & BBC sideband\\
IF\_CONV\_CHAIN & Int & & & The IF conversion chain \\
{\it (RECEIVER\_ID)}    & Int & & & Receiver id.\\
FREQ\_GROUP & Int & & & Frequency group\\
FREQ\_GROUP\_NAME & String & & & Freq. group name \\
{\it (DOPPLER\_ID)} & Int & & & Doppler id.\\
{\it (ASSOC\_SPW\_ID)} & Int(*) & & & Associated spw\_id.\\
{\it (ASSOC\_NATURE)} & String(*) & & & Nature of association\\
\defline{Flags}
FLAG\_ROW & Bool & & &\\

} \begin{description}

 \item[Notes:] This table describes properties for each defined
spectral window. A spectral window is both a frequency label for the
associated DATA array in MAIN, but also represents a generic frequency
conversion chain that shares joint physical properties and makes sense
to calibrate as a single entity.

\item[NUM\_CHAN] Number of spectral channels.

\item[NAME] Spectral window name; user specified.

\item[REF\_FREQUENCY] The reference frequency. A frequency
representative of this spectral window, usually the sky frequency
corresponding to the DC edge of the baseband. Used by the calibration
system if a fixed scaling frequency is required or in algorithms to
identify the observing band.

\item[CHAN\_FREQ] Center frequencies for each channel in the data
matrix. These can be frequency-dependent, to accommodate instruments
such as acousto-optical spectrometers. Note that the channel
frequencies may be in ascending or descending frequency order.

\item[CHAN\_WIDTH] Nomical channel width of each spectral
channel. Although these can be derived from CHAN\_FREQ by
differencing, it is more efficient to keep a separate reference to
this information.

\item[MEAS\_FREQ\_REF] Frequency Measure reference for
CHAN\_FREQ. This allows a row-based reference for this column in order to
optimize the choice of Measure reference when Doppler tracking is
used. Modified only by the MS access code.

\item[EFFECTIVE\_BW] The effective noise bandwidth of each spectral channel.

\item[RESOLUTION] The effective spectral resolution of each channel.

\item[TOTAL\_BANDWIDTH] The total bandwidth for this spectral window.

\item[NET\_SIDEBAND] The net sideband for this spectral window.

\item[BBC\_NO] The baseband converter number, if applicable.

\item[BBC\_SIDEBAND] The baseband converter sideband, is applicable.

\item[IF\_CONV\_CHAIN] Identification of the electronic signal
path for the case of multiple (simultaneous) IFs. (e.g. VLA: AC=0,
BD=1, ATCA: Freq1=0, Freq2=1)

\item[RECEIVER\_ID] Index used to identify the receiver associated
with the spectral window. Further state information is planned to be
stored in a RECEIVER sub-table.

\item[FREQ\_GROUP] The frequency group to which the spectral window belongs.
This is used to associate spectral windows for joint calibration purposes.

\item[FREQ\_GROUP\_NAME] The frequency group name; user specified.

\item[DOPPLER\_ID] The Doppler identifier defining frame information
for this spectral window.

\item[ASSOC\_SPW\_ID] Associated spectral windows, which are related
in some fashion (e.g. "channel-zero").

\item[ASSOC\_NATURE] Nature of the association for ASSOC\_SPW\_ID;
reserved keywords are ("CHANNEL-ZERO" - channel zero;
"EQUAL-FREQUENCY" - same frequency labels; "SUBSET" - narrow-band
subset).

\item[FLAG\_ROW] True if the row does not contain valid data.

\end{description}

\definetable{STATE: State information}{
\defline{\bf Columns}
\defline{\em Data}  
SIG   &     Bool  &           &        & Signal \\
REF   &     Bool  &           &        & Reference \\
CAL   &     Double &   K      &        & Noise calibration \\
LOAD  &     Double &   K      &        & Load temperature\\
SUB\_SCAN &  Int   &          &        & Sub-scan number\\
OBS\_MODE &  String &         &        & Observing mode\\
\defline{\em Flags}
FLAG\_ROW & Bool &            &        & Row flag\\
}
\begin{description}

\item[Notes:] This table defines the state parameters for a particular
data record as they refer to external loads, calibration sources or
references, and also characterizes the observing mode of the data
record, as an aid to defining the scheduling heuristics. It is indexed
directly via STATE\_ID in MAIN.

\item[SIG] True if the source signal is being observed.

\item[REF] True for a reference phase.

\item[CAL] Noise calibration temperature (zero if not added).

\item[LOAD] Load temperature (zero if no load).

\item[SUB\_SCAN] Sub-scan number $(\geq 0)$, relative to the
SCAN\_NUMBER in MAIN. Used to identify observing sequences.

\item[OBS\_MODE] Observing mode; defined by a set of reserved keywords
characterizing the current observing mode (e.g. "OFF-SPECTRUM"). Used
to define the schedule strategy.

\item[FLAG\_ROW] True if the row does not contain valid data. Does not
imply flagging in MAIN.

\end{description}

\definetable{SYSCAL: System calibration}{
\defline{\bf Columns}
\defline{\em Key}
ANTENNA\_ID &     Int &    &      & Antenna id\\
FEED\_ID &        Int &    &      & Feed id\\
SPECTRAL\_WINDOW\_ID &   Int &    &      & Spectral window id\\
TIME      &   Double  &      s  & EPOCH & Midpoint of time for which this set of parameters is accurate\\
INTERVAL &       Double  &      s     &   & Interval\\
\defline{\em Data}  
{\it (PHASE\_DIFF)} &  Float & rad &    & Phase difference between receptor 0 and receptor 1\\
{\it (TCAL)} &       Float (\nr)  &      K &  & Calibration temp\\
{\it (TRX)}  &       Float (\nr)  &      K &  & Receiver temperature\\
{\it (TSKY)} &       Float (\nr)  &      K &  & Sky temperature\\
{\it (TSYS)} &       Float (\nr)  &      K &  & System temp\\
{\it (TANT)} &       Float (\nr)  &      K &  & Antenna temperature\\
{\it (TANT\_TSYS)} & Float(\nr) &  & & ${{T_{ant}}\over{T_{sys}}}$\\
{\it (TCAL\_SPECTRUM)} &       Float (\nr, \nf)  &      K &  & Calibration temp\\
{\it (TRX\_SPECTRUM)}  &       Float (\nr, \nf)  &      K &  & Receiver temperature\\
{\it (TSKY\_SPECTRUM)}  &       Float (\nr, \nf)  &      K &  & Sky temperature spectrum\\
{\it (TSYS\_SPECTRUM)} &       Float (\nr, \nf)  &      K &  & System temp\\
{\it (TANT\_SPECTRUM)} &       Float (\nr, \nf)  &      K &  & Antenna temperature spectrum\\
{\it (TANT\_TSYS\_SPECTRUM)} & Float (\nr,\nf) &  & & ${{T_{ant}}\over{T_{sys}}}$ spectrum\\
\defline{\em Flags}
{\it (PHASE\_DIFF\_FLAG)} &   Bool  &    &      & Flag for PHASE\_DIFF\\
{\it (TCAL\_FLAG)} &  Bool &    &      & Flag for TCAL\\
{\it (TRX\_FLAG)} &   Bool  &    &      & Flag for TRX\\
{\it (TSKY\_FLAG)} & Bool &      &      & Flag for TSKY\\
{\it (TSYS\_FLAG)} &  Bool &    &      & Flag for TSYS\\
{\it (TANT\_FLAG)} & Bool & & & Flag for TANT\\
{\it (TANT\_TSYS\_FLAG)} & Bool & & & Flag for ${{T_{ant}}\over{T_{sys}}}$\\
}
\begin{description}

\item[Notes:] This table contains time-variable calibration
measurements for each antenna, as indexed on feed and spectral window.
Note that \nr = number of receptors, and \nf = number of
frequency channels.

\item[ANTENNA\_ID] Antenna identifier, as indexed by ANTENNA$n$ in MAIN.

\item[FEED\_ID] Feed identifier, as indexed by FEED$n$ in MAIN.

\item[SPECTRAL\_WINDOW\_ID] Spectral window identifier.

\item[TIME] Mid-point of the time interval for which the data in this
row are valid. Required to use the same TIME Measure reference as that
in MAIN.

\item[INTERVAL] Time interval.

\item[PHASE\_DIFF] Phase difference between receptor 0 and receptor 1.

\item[TCAL] Calibration temperature.

\item[TRX] Receiver temperature.

\item[TSKY] Sky temperature.

\item[TSYS] System temperature.

\item[TANT] Antenna temperature.

\item[TANT\_TSYS] Antenna temperature over system temperature.

\item[TCAL\_SPECTRUM] Calibration temperature spectrum.

\item[TRX\_SPECTRUM] Receiver temperature spectrum.

\item[TSKY\_SPECTRUM] Sky temperature spectrum.

\item[TSYS\_SPECTRUM] System temperature spectrum.

\item[TANT\_SPECTRUM] Antenna temperature spectrum.

\item[TANT\_TSYS\_SPECTRUM] Antenna temperature over system
temperature spectrum.

\item[PHASE\_DIFF\_FLAG] True if PHASE\_DIFF flagged.

\item[TCAL\_FLAG] True if TCAL flagged.

\item[TRX\_FLAG] True if TRX flagged.

\item[TSKY\_FLAG] True if TSKY flagged.

\item[TSYS\_FLAG] True if TSYS flagged.

\item[TANT\_FLAG] True if TANT flagged.

\item[TANT\_TSYS\_FLAG] True if TANT\_TSYS flagged.

\end{description}


\definetable{WEATHER: weather station information}{
\defline{\bf Columns}
\defline{\em Key}
ANTENNA\_ID &        Int &    &      & Antenna number\\
TIME &  Double  &   s  & EPOCH  & Mid-point of interval\\ 
INTERVAL &       Double &        s  &      & Interval over which data is relevant\\
\defline{\em Data}  
{\it (H2O)} &        Float    &     $m^{-2}$  &      & Average column density of water\\
{\it (IONOS\_ELECTRON)} & Float & $m^{-2}$    &      & Average column density of electrons\\
{\it (PRESSURE)} &       Float & hPa  &      & Ambient atmospheric pressure\\
{\it (REL\_HUMIDITY)} &   Float &    &      & Ambient relative humidity\\
{\it (TEMPERATURE)} &    Float &  K  &     & Ambient air temperature for an antenna\\
{\it (DEW\_POINT)} & Float & K & & Dew point \\
{\it (WIND\_DIRECTION)} & Float  &       rad &      & Average wind direction\\
{\it (WIND\_SPEED)} &  Float &  m/s  &    & Average wind speed \\
\defline{\em Flags}
{\it (H2O\_FLAG)} & Bool & & & Flag for H2O \\
{\it (IONOS\_ELECTRON\_FLAG)} & Bool & & & Flag for IONOS\_ELECTRON \\
{\it (PRESSURE\_FLAG)} & Bool & & & Flag for PRESSURE \\
{\it (REL\_HUMIDITY\_FLAG)} & Bool & & & Flag for REL\_HUMIDITY \\
{\it (TEMPERATURE\_FLAG)} & Bool & & & Flag for TEMPERATURE \\
{\it (DEW\_POINT\_FLAG)} & Bool & & & Flag for DEW\_POINT \\
{\it (WIND\_DIRECTION\_FLAG)} & Bool & & & Flag for WIND\_DIRECTION \\
{\it (WIND\_SPEED\_FLAG)} & Bool & & & Flag for WIND\_SPEED \\
}
\begin{description}

\item[Notes:] This table contains mean external atmosphere and weather
information.

\item[ANTENNA\_ID] Antenna identifier, as indexed by ANTENNA$n$ from MAIN.

\item[TIME] Mid-point of the time interval over which the data in the
row are valid. Required to use the same TIME Measure reference as in
MAIN.

\item[INTERVAL] Time interval.

\item[H2O] Average column density of water.

\item[IONOS\_ELECTRON] Average column density of electrons.

\item[PRESSURE] Ambient atmospheric pressure.

\item[REL\_HUMIDITY] Ambient relative humidity.

\item[TEMPERATURE] Ambient air temperature.

\item[DEW\_POINT] Dew point temperature.

\item[WIND\_DIRECTION] Average wind direction.

\item[WIND\_SPEED] Average wind speed.

\item[H2O\_FLAG] Flag for H2O.

\item[IONOS\_ELECTRON\_FLAG] Flag for IONOS\_ELECTRON.

\item[PRESSURE\_FLAG] Flag for PRESSURE.

\item[REL\_HUMIDITY\_FLAG] Flag for REL\_HUMIDITY.

\item[TEMPERATURE\_FLAG] Flag for TEMPERATURE.

\item[DEW\_POINT\_FLAG] Flag for DEW\_POINT.

\item[WIND\_DIRECTION\_FLAG] Flag for DEW\_POINT.

\item[WIND\_SPEED\_FLAG] Flag for DEW\_POINT.

\end{description}

\section{Correction and additions}

\begin{description}

\item{{\bf 2000/08/14:} Add expanded description of Doppler tracking to
Section 4.0.8; correctly identify SOURCE sub-table as optional; expand
definition of SPECTRAL\_WINDOW\_ID=-1 in the description of the SOURCE
sub-table; expand the definition of the REF\_FREQUENCY in the
SPECTRAL\_WINDOW sub-table description.}
\item{{\bf 2000/08/28:} Make POSITION field in SOURCE table optional.}
\item{{\bf 2000/09/12:} Add required VELDEF column in DOPPLER table.}

\end{description}

\section{References}

\noindent Date, C.J., 1986, Introduction to database systems, (Addison-Wesley).

\noindent Flatters, C., 1998, AIPS Memo. 102.

\noindent Noordam, J., 1996, aips++ Note 185.

\noindent Wieringa, M.H., and Cornwell, T., 1996, aips++ Note 191.

\end{document}

